---
title: <font size="5">**OPTVR -- Walking speed change detection**</font> 
author: <br> <font size="4"> Pawe³ Motyka (University of Warsaw) </font> <br>  *pawel.motyka@psych.uw.edu.pl* 
date: <font size="3"> January 2020  </font>
output: html_document
chunk_output_type: console

--- 

&nbsp;

<font size="4">
**List of sections**:

1. Load required packages and process the data [S1](#S1)
2.  Derive signal detection theory (SDT) measures [S2](#S2)
3.  Plot hit rates and false alarms [S3](#S3)
4.  Compare SDT-measures at different conditions [S4](#S4)
5.  Explore associations bewteen SDT-measures and confidence & anthropometric data [S5](#S5)

<a name="S1"></a>
&nbsp;

#####**1. Load required packages and proprocess the data** 

```{r, message=FALSE, warning= FALSE}

library(dplyr)
library(ggplot2)
library(colorspace)
library(scales)
library(effsize)
library(tidyr)
library(psych)
library(cowplot)
library(ggpubr)
library(plyr)
library(readr)
library(Rmisc)
library(devtools)
library(gghalves)
library(coin)
library(here)

#data directory & data reading
data_dir <- paste0(here(),"/_data")
setwd(data_dir)

# read walking speed change detection data
df <- read.table("OPTVR_walking_speed_change_detection_data", sep = "\t", header = T)

# read confidence ratings rata
df_conf <- read.table("OPTVR_walking_speed_change_detection_conf_data", sep = "\t", header = T)

# read demographic and anthropometric data
df_sub_data <- read.table("OPTVR_determination_of_optic_flow_speed_data", sep = "\t", header = T)
df_sub <- df_sub_data[,c(1:5)]
df_sub <- df_sub[complete.cases(df_sub),]
df_sub$Gender <- plyr::revalue(df_sub$Gender, replace = c("W " = "W")) #correction

# classify as a hit in case of the stimulus present and "yes" response
df$r[df$Change == "Yes" & df$Resp == "y"] <- 'hit' 

# classify as a miss in case of the stimulus present and "no" response
df$r[df$Change == "Yes" & df$Resp == "n"] <- 'miss' 

# classify as a false alarm in case of the catch trial and "yes" response
df$r[df$Change == "no" & df$Resp == "y"] <- 'fa' 

# classify as a correct rejection in case of the catch trial and "no" response
df$r[df$Change == "no" & df$Resp == "n"] <- 'cr' 

# save as factor
df$r <- factor(df$r)

# remove training blocks
df <- df[df$Block_num != 99,]
df <- df[df$Block_num != 98,]

```


<a name="S2"></a>
&nbsp;

#####**2. Derive signal detection theory (SDT) measures** 

```{r}

# data subjects level
dfs <- data.frame(ID = integer(0),
                   hit_rate = numeric (0), #hit rate
                   fa_rate = numeric(0), # false alarms rate
                   s = numeric(0), # sensitivity (d')
                   c = numeric (0), # criterion (c)
                   hit_rate_acc = numeric (0), #acceleration blocks
                   hit_rate_dec = numeric (0), # deceleration blocks
                   fa_rate_acc = numeric(0),
                   fa_rate_dec = numeric(0),
                   s_acc = numeric(0), 
                   s_dec = numeric(0),
                   c_acc = numeric (0),
                   c_dec = numeric (0)) 
  
  
  for (p in unique(df$ID)) {     # LOOP PARTICIPANTS (p)

      #calculate hit rate
      hit_rate <- (length(df$r[df$r =="hit" & df$ID == p])+0.5)/(length(df$Change[df$ID == p & df$Change == "Yes"])+1)

      # calculate false alarm rate
      FA_rate <- (length(df$r[df$r =="fa" & df$ID == p])+0.5)/(length(df$Change[df$ID == p & df$Change == "no"])+1)
    
      # calculate sensitivity (d') [d' = Z(hit_rate) - Z(False_alarms_rate)] 
      s <- qnorm(hit_rate) - qnorm(FA_rate) 
      
      # calculate criterion (c) [c = - [Z(hit_rate) + Z(False_ alarms_rate)]/2] 
      c <- -(qnorm(hit_rate)+qnorm(FA_rate))/2 
      
    # calculate hit rate - acceleration
      hit_rate_acc <- (length(df$r[df$r =="hit" & df$ID == p & df$Block_type == "speed_up"])+0.5)/(length(df$Change[df$ID == p & df$Change == "Yes" & df$Block_type == "speed_up"])+1)
      
      # calculate hit rate - deceleration
       hit_rate_dec <- (length(df$r[df$r =="hit" & df$ID == p & df$Block_type == "speed_down"])+0.5)/(length(df$Change[df$ID == p & df$Change == "Yes" & df$Block_type == "speed_down"])+1)
      
      # false alarms - acceleration
      FA_rate_acc <- (length(df$r[df$r =="fa" & df$ID == p & df$Block_type == "speed_up"])+0.5)/(length(df$Change[df$ID == p & df$Change == "no" & df$Block_type == "speed_up"])+1)
      
      # false alarms - deceleration
       FA_rate_dec <- (length(df$r[df$r =="fa" & df$ID == p & df$Block_type == "speed_down"])+0.5)/(length(df$Change[df$ID == p & df$Change == "no" & df$Block_type == "speed_down"])+1)
      
      # Sensitivity (d') [d' = Z(hit_rate) - Z(False_alarms_rate)] 
      s_acc <- qnorm(hit_rate_acc) - qnorm(FA_rate_acc)
      s_dec <- qnorm(hit_rate_dec) - qnorm(FA_rate_dec) 
      
      # Criterion (c) [c = - [Z(hit_rate) + Z(False_ alarms_rate)]/2] 
      c_acc <- -(qnorm(hit_rate_acc)+qnorm(FA_rate_acc))/2 
      c_dec <- -(qnorm(hit_rate_dec)+qnorm(FA_rate_dec))/2 
     
      # create a dfframe with sensitivity and criterion measures
      dfs[nrow(dfs)+1,] <- c(p, hit_rate, FA_rate, s, c, hit_rate_acc, hit_rate_dec, FA_rate_acc, FA_rate_dec, s_acc,s_dec,c_acc, c_dec)
      
  } # END: LOOP PARTICIPANTS (p)

# hit and fa rates in %
dfs$hit_rate <- dfs$hit_rate * 100
dfs$hit_rate_acc <- dfs$hit_rate_acc * 100
dfs$hit_rate_dec <- dfs$hit_rate_dec * 100
dfs$fa_rate <- dfs$fa_rate * 100
dfs$fa_rate_acc <- dfs$fa_rate_acc * 100
dfs$fa_rate_dec <- dfs$fa_rate_dec * 100

```

<a name="S3"></a>
&nbsp;

#####**3. Plot hit rates and false alarms** 

```{r}

### ACCELERATION 
#hit rate & fa rate
describe(dfs$hit_rate_acc)
describe(dfs$fa_rate_acc)

## DECELERATION
#hit rate & fa rate
describe(dfs$hit_rate_dec)
describe(dfs$fa_rate_dec)

### plot

d <- select(dfs, ID, fa_rate_acc, fa_rate_dec, hit_rate_acc, hit_rate_dec)

d <- gather(d, key = variable, value = value, -"ID")

d$condition <- c()
d$measure <- c()
  
for (r in c(1:nrow(d))) {
d$condition[r] <- ifelse(d$variable[r] == "hit_rate_acc" | d$variable[r] == "fa_rate_acc", "acc", "dec")
d$variable[r] <- ifelse(d$variable[r] == "hit_rate_acc" | d$variable[r] == "hit_rate_dec", "hit_rate", "fa_rate")
}

d <- pivot_wider(d, names_from = c(variable), values_from = c(value))


bold.text <- element_text(face = "bold", color = "black")

d$jitter_fa <- runif(nrow(d), min = -1.5, max=1.5)
d$jitter_hits <- runif(nrow(d), min = -1.5, max= 1.5)

d$fa_rate_jit <- d$fa_rate + d$jitter_fa
d$hit_rate_jit <- d$hit_rate + d$jitter_hits

col1 <- darken("red3", amount = 0.1)
col2 <- "dodgerblue4"

# Create scatter plot
fig <- ggscatter(d, x = "fa_rate_jit", y = "hit_rate_jit", color = "condition", alpha = 0.4, shape = "condition", size = "condition") + 
border() + 
labs(x = "False alarms (%)", y = "Hit rate (%)") +
scale_shape_manual(values = c('acc' = 19, 'dec' = 17)) +
scale_size_manual(values = c('acc' = 2.2, 'dec' = 2)) +
scale_color_manual(values = c('acc' = darken(col1, amount = 0.4), 'dec' = darken(col2, amount = 0.4))) + 
scale_y_continuous(limits= c(0, 100), breaks = c(0,25,50,75,100), expand = c(0.01,0.01)) + 
scale_x_continuous(limits= c(0, 100), breaks = c(0,25,50,75,100), expand = c(0.01,0.01)) +
theme(plot.margin=unit(c(7,10,7,7),"pt"), axis.text = element_text(face="bold",size=14),axis.title = element_text(face="bold",size=14, colour = "black"), axis.title.y = element_text(margin = margin(t = 0, r = 3, b = 0, l = 0)), axis.title.x =   element_text(margin = margin(t = 9.5, r = 0, b = 0, l = 0))) + geom_abline(slope=1, intercept= 0, colour = scales::alpha("gray40",0.5), lwd = 1, linetype = "dashed")


# Marginal densities for x and y axes
xplot <- ggdensity(d, "fa_rate", fill = "condition") + scale_x_continuous(limits= c(0, 100)) + scale_fill_manual(values = c('acc' = darken(col1, amount = 0.1), 'dec' = darken(col2, amount = 0.3))) +
annotate("segment", x = mean(d$fa_rate[d$condition == "acc"]), xend = mean(d$fa_rate[d$condition == "acc"]), y = 0.0, yend = 0.0205, size = 1.5, colour = scales::alpha(darken(col1, amount = 0.3),0.6)) + annotate("segment", x = mean(d$fa_rate[d$condition == "dec"]), xend = mean(d$fa_rate[d$condition == "dec"]), y = 0.0, yend = 0.0248, size = 1.5, colour = scales::alpha(darken(col2, amount = 0.5),0.6))

yplot <- ggdensity(d, "hit_rate", fill = "condition") +  scale_x_continuous(limits= c(0, 100)) + scale_fill_manual(values = c('acc' = darken(col1, amount =0.1), 'dec' = darken(col2, amount = 0.3))) +
annotate("segment", x = mean(d$hit_rate[d$condition == "acc"]), xend = mean(d$hit_rate[d$condition == "acc"]), y = 0.0, yend = 0.0198, size = 1.5, colour = scales::alpha(darken(col1, amount = 0.3),0.6)) + annotate("segment", x = mean(d$hit_rate[d$condition == "dec"]), xend = mean(d$hit_rate[d$condition == "dec"]), y = 0.0, yend = 0.029, size = 1.5, colour = scales::alpha(darken(col2, amount = 0.5),0.6)) + rotate()

# reduce plots
fig <- fig + rremove("legend")

yplot <- yplot + clean_theme() + rremove("legend") + theme(plot.margin=unit(c(5,0,0,0),"pt")) 


xplot <- xplot + clean_theme() + rremove("legend") + 
  theme(plot.margin=unit(c(0,5,0,0),"pt")) 

# arrange plots
plot_grid(xplot, NULL, fig, yplot, nrow = 2, ncol = 2, align = "hv", 
      rel_widths = c(3, 1), rel_heights = c(1, 3))
#export size 5 x 5.5

```


<a name="S4"></a>
&nbsp;

#####**4. Compare SDT-measures at different conditions**

```{r}

### Compare SDT-measures between tasks 

## Sensitivity

# summary statistics
describe(dfs$s_acc)
describe(dfs$s_dec)

# test distributions
shapiro.test(dfs$s_acc)
shapiro.test(dfs$s_dec)

# test differences
test <- wilcoxsign_test(dfs$s_acc ~ dfs$s_dec, paired = T, distribution = "exact")
test

## Criterion

# summary statistics
describe(dfs$c_acc)
describe(dfs$c_dec)

# test distributions
shapiro.test(dfs$c_acc)
shapiro.test(dfs$c_dec)

# test differences
test <- wilcoxsign_test(dfs$c_acc ~ dfs$c_dec, paired = T, distribution = "exact")
test

# effect size
test@statistic@teststatistic / sqrt(nrow(dfs))

# s_both = c(dfs$s_acc, dfs$s_dec)
# gp = c(rep(1,46), rep(2,46))
# 
# boxplot(s_both ~ gp, col=c("brown3", "darkorchid3"), pch=19, ylab = "Sensitivity", line = 2.5, cex.lab= 1.6)


## Plot differences in Sensitivity

y_lim_min = -0.5
y_lim_max = 3

d <- select(dfs, ID, s_acc, s_dec)

d <- gather(d, key = x, value = y, -"ID")

d$x[d$x == "s_acc"] <- 1
d$x[d$x == "s_dec"] <- 2
d$x <- as.numeric(d$x)
d$xj <- jitter(d$x, amount=.04)

fig <- ggplot(data = d, aes(y = y)) +
  
   #Add geom_() objects
   geom_point(data = d %>% filter(x =="1"), aes(x = xj), color =  darken(col1, amount =0.1), size = 1.5, alpha = .5) +
   geom_point(data = d %>% filter(x =="2"), aes(x = xj), color = darken(col2, amount = 0.3), size = 1.5, alpha = .5) +
   geom_line(aes(x = xj, group = ID), color = 'lightgray', alpha = .5) +
  
   geom_half_boxplot( data = d %>% filter(x=="1"), aes(x=x, y = y), position = position_nudge(x = -.25), side = "r",outlier.shape = NA, center = TRUE, errorbar.draw = TRUE, width = .3, fill =  darken(col1, amount =0.1), alpha = .5) +
   
   geom_half_boxplot(data = d %>% filter(x=="2"), aes(x=x, y = y), position = position_nudge(x = .15), side = "r",outlier.shape = NA, center = TRUE, errorbar.draw = TRUE, width = .3, fill = darken(col2, amount = 0.3), alpha = .5) +
  
   geom_half_violin(data = d %>% filter(x=="1"),aes(x = x, y = y), position = position_nudge(x = 1.4), side = "r", fill =  darken(col1, amount =0.1), alpha = .5, trim = FALSE) +
 
   geom_half_violin(data = d %>% filter(x=="2"),aes(x = x, y = y), position = position_nudge(x = .4), side = "r", fill = darken(col2, amount = 0.3), alpha = .5, trim = FALSE) +
  
  #Define additional settings
   scale_x_continuous(breaks=c(1,2), labels=c("Acceleration", "Deceleration"), limits=c(0.5,3)) +
   xlab("Condition") + ylab("Sensitivity (d')") +
   theme_classic()+
   coord_cartesian(ylim=c(y_lim_min, y_lim_max)) + theme(plot.margin=unit(c(7,10,7,7),"pt"), axis.text.x = element_text(size=14, colour = "black"), axis.text.y = element_text(vjust = 0.2, face="bold",size=15, colour = "black"), axis.title.y = element_text(margin = margin(t = 0, r = 8, b = 0, l = 0), face="bold",size=17, colour = "black"), axis.title.x = element_text(hjust=0.37, margin = margin(t = 7, r = 0, b = 0, l = 0), face="bold",size= 14, colour = "black"))

fig
# export 3.5 x 4.5 each

## Plot differences in Criterion

y_lim_min = -1.4
y_lim_max = 2

d <- select(dfs, ID, c_acc, c_dec)

d <- gather(d, key = x, value = y, -"ID")

d$x[d$x == "c_acc"] <- 1
d$x[d$x == "c_dec"] <- 2

d$x <- as.numeric(d$x)
d$xj <- jitter(d$x, amount=.04)

fig <- ggplot(data = d, aes(y = y)) +
  
   #Add geom_() objects
   geom_point(data = d %>% filter(x =="1"), aes(x = xj), color =  darken(col1, amount =0.1), size = 1.5, alpha = .5) +
   geom_point(data = d %>% filter(x =="2"), aes(x = xj), color = darken(col2, amount = 0.3), size = 1.5, alpha = .5) +
   geom_line(aes(x = xj, group = ID), color = 'lightgray', alpha = .5) +
  
   geom_half_boxplot( data = d %>% filter(x=="1"), aes(x=x, y = y), position = position_nudge(x = -.25), side = "r",outlier.shape = NA, center = TRUE, errorbar.draw = TRUE, width = .3, fill =  darken(col1, amount =0.1), alpha = .5) +
   
   geom_half_boxplot(data = d %>% filter(x=="2"), aes(x=x, y = y), position = position_nudge(x = .15), side = "r",outlier.shape = NA, center = TRUE, errorbar.draw = TRUE, width = .3, fill = darken(col2, amount = 0.3), alpha = .5) +
  
   geom_half_violin(data = d %>% filter(x=="1"),aes(x = x, y = y), position = position_nudge(x = 1.4), side = "r", fill =  darken(col1, amount =0.1), alpha = .5, trim = FALSE) +
 
   geom_half_violin(data = d %>% filter(x=="2"),aes(x = x, y = y), position = position_nudge(x = .4), side = "r", fill = darken(col2, amount = 0.3), alpha = .5, trim = FALSE) +
  
  #Define additional settings
   scale_x_continuous(breaks=c(1,2), labels=c("Acceleration", "Deceleration"), limits=c(0.5,3)) +
   xlab("Condition") + ylab("Criterion (C)") +
   theme_classic()+
   coord_cartesian(ylim=c(y_lim_min, y_lim_max)) + theme(plot.margin=unit(c(7,10,7,7),"pt"), axis.text.x = element_text(size=14, colour = "black"), axis.text.y = element_text(face="bold",size=15, colour = "black"), axis.title.y = element_text(margin = margin(t = 0, r = 1, b = 0, l = 0), face="bold",size=17.5, colour = "black"), axis.title.x = element_text(hjust=0.37, margin = margin(t = 7, r = 0, b = 0, l = 0), face="bold",size= 14, colour = "black"))

fig
# export 3.5 x 4.5 each


## Sensitivity 
shapiro.test(dfs$s_acc)
shapiro.test(dfs$s_dec)
cor.test(dfs$s_acc, dfs$s_dec, method = "spearman", exact = F)
cor.test(dfs$s_acc, dfs$s_dec, method = "pearson")

# Plot
bold.text <- element_text(face = "bold", color = "black")
Figure_sensitivity <- ggplot(data = dfs, aes(x = s_acc, y = s_dec)) + geom_smooth(col = "grey8", method = "lm", level=0.95, alpha = 0.5) +  geom_point(col = "grey 15", alpha = 0.5, size = 3, shape = 19) +   labs(x = "d' - acceleration", y = "d' - deceleration") + scale_fill_manual() + theme_classic() +  theme(axis.text=element_text(size=15, colour = "black", face="bold"),  axis.title=element_text(size=17), axis.title.y = element_text(margin = margin(t = 0, r = 3, b = 0, l = 0)), axis.title.x =   element_text(margin = margin(t = 5, r = 0, b = 0, l = 0))) + theme(axis.title = bold.text)
Figure_sensitivity

## Criterion
shapiro.test(dfs$c_acc)
shapiro.test(dfs$c_dec)
cor.test(dfs$c_acc, dfs$c_dec, method = "spearman", exact = F)

# Plot
bold.text <- element_text(face = "bold", color = "black")
Figure_criterion <- ggplot(data = dfs, aes(x = c_acc, y = c_dec)) +  geom_smooth(col = "grey8", method = "lm", level=0.95, alpha = 0.5) + geom_point(col = "grey 15", alpha = 0.5, size = 3, shape = 19) + labs(x = "C - acceleration", y = "C - deceleration") +  scale_fill_manual() +  theme_classic() +  theme(axis.text=element_text(size=15, colour = "black", face="bold"),  axis.title=element_text(size=17), axis.title.y = element_text(margin = margin(t = 0, r = 3, b = 0, l = 0)), axis.title.x =   element_text(margin = margin(t = 5, r = 0, b = 0, l = 0))) + theme(axis.title = bold.text)
Figure_criterion
# 3.5 x 4.1

```



<a name="S5"></a>
&nbsp;

#####**5. Explore associations between SDT-measures and other variables** 


```{r}

### Merge with confidence ratings
dfs <- merge(dfs, df_conf, by = "ID")
dfs<- plyr::rename(dfs, replace = c("conf_speed_up" = "conf_acc"))
dfs<- plyr::rename(dfs, replace = c("conf_speed_down" = "conf_dec"))
df_sub <- unique(df_sub)
dfs <- merge(dfs, df_sub, by = "ID")
dfs$conf_acc_dec_mean <- (dfs$conf_acc + dfs$conf_dec)/2

# summary statistics 
describe(dfs$conf_dec)
describe(dfs$conf_acc)

# test distributions
shapiro.test(dfs$conf_dec)
shapiro.test(dfs$conf_acc)

# test differences
test <- wilcoxsign_test(dfs$conf_dec ~ dfs$conf_acc, paired = T, distribution = "exact")
test

# effect size
test@statistic@teststatistic / sqrt(nrow(dfs))

# correlation between confidence ratings
cor.test(dfs$conf_acc, dfs$conf_dec, method = "spearman", exact = F)
#scatter.smooth(dfs$conf_acc, dfs$conf_dec)

### correlation between SDT-measures and confidence ratings

cor.test(dfs$s, dfs$conf_acc_dec_mean, method = "spearman", exact = F)
#scatter.smooth(dfs$s, dfs$conf_acc_dec_mean)

cor.test(dfs$s_acc, dfs$conf_acc, method = "spearman", exact = F)
#scatter.smooth(dfs$s_acc, dfs$conf_acc)

cor.test(dfs$s_dec, dfs$conf_dec, method = "spearman", exact = F)
#scatter.smooth(dfs$s_dec, dfs$conf_dec)

cor.test(dfs$c_acc, dfs$conf_acc, method = "spearman", exact = F)
#scatter.smooth(dfs$c_acc, dfs$conf_acc)

cor.test(dfs$c_dec, dfs$conf_dec, method = "spearman", exact = F)
#scatter.smooth(dfs$c_dec, dfs$conf_dec)


## correlation between SDT-measures and anthropometric variables
dfs$bmi <- dfs$Weight / ((dfs$Height)/100)^2
describe(dfs$bmi)

shapiro.test(dfs$bmi)
shapiro.test(dfs$s)

cor.test(dfs$bmi, dfs$s, method = "spearman", exact = F)
#scatter.smooth(dfs$bmi, dfs$s)

cor.test(dfs$Height, dfs$s, method = "spearman", exact = F)
#scatter.smooth(dfs$Height, dfs$s)

cor.test(dfs$Weight, dfs$s, method = "spearman", exact = F)
#scatter.smooth(dfs$Weight, dfs$s)

## BMI between genders
shapiro.test(dfs$bmi[dfs$Gender == "W"])
shapiro.test(dfs$bmi[dfs$Gender == "M"])
t.test(dfs$bmi[dfs$Gender == "M"], dfs$bmi[dfs$Gender == "W"], paired = T)
effsize::cohen.d(dfs$bmi[dfs$Gender == "M"], dfs$bmi[dfs$Gender == "W"], paired = T)
describe(dfs$bmi[dfs$Gender == "M"])
describe(dfs$bmi[dfs$Gender == "W"])

# correlations within sub-samples
shapiro.test(dfs$s[dfs$Gender == "M"])
cor.test(dfs$bmi[dfs$Gender == "M"], dfs$s[dfs$Gender == "M"], method = "spearman", exact = F)

shapiro.test(dfs$s[dfs$Gender == "W"])
cor.test(dfs$bmi[dfs$Gender == "W"], dfs$s[dfs$Gender == "W"], method = "spearman", exact = F)

## Overall sensitivity between genders
# test distributions
shapiro.test(dfs$s[dfs$Gender == "W"])
shapiro.test(dfs$s[dfs$Gender == "M"])

# statistical summary
describe(dfs$s[dfs$Gender == "W"])
describe(dfs$s[dfs$Gender == "M"])

# test differences
wilcox.test(dfs$s[dfs$Gender == "W"], dfs$s[dfs$Gender == "M"], exact = F, paired = F)

# sample data
describe(dfs$Age)
length(dfs$ID[dfs$Gender == "W"])
length(dfs$ID[dfs$Gender == "M"])

#write.table(dfs, file = "OPTVR_walking_speed_change_detection_output", sep = "\t")

```


