---
title: <font size="5">**OPTVR -- Determination of optic flow speeds**</font> 
author: <br> <font size="4"> Pawe³ Motyka (University of Warsaw) </font> <br>  *pawel.motyka@psych.uw.edu.pl* 
date: <font size="3"> January 2020  </font>
output: html_document
chunk_output_type: console

--- 

&nbsp;

<font size="4">

**List of sections**:

1. Load required packages and process the data [S1](#S1)
2. Visualize overall detection performance [S2](#S2)
3. Compare matched optic flow speeds at different conditions [S3](#S3)
4. Explore associations bewteen optic flow estimates and confidence & anthropometric & proprioceptive data [S4](#S4)

<a name="S1"></a>
&nbsp;

#####**1. Load the required packages and process the data** 

```{r, message=FALSE, warning= FALSE}

# libraries
require(dplyr)
require(ggplot2)
require(effsize)
require(scales)
require(effsize)
require(psych)
require(here)
require(coin)
require(colorspace)
require(tidyr)

# data directory
data_dir <- paste0(here::here(),"/_data")
setwd(data_dir)

# read optic flow speed data
df <- read.table("OPTVR_determination_of_optic_flow_speed_data", header = TRUE, sep = "\t", fill = TRUE, stringsAsFactors = FALSE)

# derive speed estimates for different conditions
d <- df %>%
  group_by(ID) %>%
  dplyr::summarize(mF = median(Estimate[Walking == "faster"]),
            mS = median(Estimate[Walking == "slower"]),
            sdF = sd(Estimate[Walking == "faster"]),
            sdS = sd(Estimate[Walking == "slower"]),
            m_starting_F = median(Estimate[Starting_speed == "F"]),
            m_starting_S = median(Estimate[Starting_speed == "S"]),
            m_red = median(Estimate[Color == "Red"]),
            m_green = median(Estimate[Color == "Green"]))


# add demographic & anthropometric data
d_static <- unique(df[c(1:5,12)])
d_static$bmi <- d_static$Weight / ((d_static$Height/100) * (d_static$Height/100))
d_static$Gender <- plyr::revalue(d_static$Gender, replace = c("W " = "W"))
d <- merge(d, d_static, by = "ID")

# standardized values (outliers detection)
d$mF_z <- (d$mF - mean(d$mF)) / sd(d$mF)
d$mS_z <- (d$mS - mean(d$mS)) / sd(d$mS)

# show outliers (more than 3SD above or below the grand average)
d$ID[d$mF_z > 3 | d$mF_z < -3]
d$ID[d$mS_z > 3 | d$mS_z < -3]

# difference between matched optic flow speeds at different walking conditions
d$diff_walking_speed <- d$mF - d$mS

# check subjects with inversed optic flow speeds (slower > faster)
d$ID[d$diff_walking_speed < 0]

# exclude one person with inversed
d <- d[d$ID != 20, ]

# difference between matched optic flow speeds at different starting (visual) speeds
d$diff_starting_speed <- d$m_starting_F - d$m_starting_S

# difference between matched optic flow speeds at different colours of visualization
d$diff_color <- d$m_green - d$m_red

# read the preprocessed walking speed change detection data
dt_walking <- read.table("OPTVR_walking_speed_change_detection_output", header = TRUE, sep = "\t", fill = TRUE, stringsAsFactors = FALSE)
dt_walking <- dt_walking %>% select(-Age,-Gender,-bmi,-Height,-Weight)

```

<a name="S2"></a>
&nbsp;

#####**2. Visualize overall detection performance** 

```{r}

# save data as vectors
mF <- d$mF
mS <- d$mS

# color
col <- darken("#d1ab02", 0.07)
col2 <- darken("#a18300", 0.1)

# generate square plotting region
par(pty="s", font.axis = 2)

# specify plotting space
plot(c(0,3),c(0,3),type="n",xlab ="", ylab="", cex.lab = 1, cex.axis = 0.8, frame.plot = F)

# add labs 
title(ylab="Estimate (m/s) - walking faster", line = 2.5, cex.lab= 1)
title(xlab="Estimate (m/s) - walking slower", line = 2.5, cex.lab= 1)

# create an identity line using a customized linear model
x<-0:3
y<-0:3
line <- data.frame(x = seq(0, 2.8, 0.1))
lines(line$x, predict(lm(y~x), line),col= scales::alpha("black", alpha = 0.5),lty= 2, lwd = 1.3)

# add line marking an accurate ratio between estimates
abline(c(0,(2.5/1.5)), col=scales::alpha("black",0.5), lty = 1, lwd = 1.3)

# plot the individual data
points(mS,mF, pch = 19, col = scales::alpha(col, alpha = 0.4), cex = 1)
#points(0.416,0.400, pch = "+", col = alpha("gray35", alpha = 1), cex = 1.9, lwd = 1.5) # values of y axis corrected to properly center "+" sign

# plot grand means
points(median(d$mS), median(d$mF), pch = 3, col = scales::alpha(col2, alpha = 1), cex = 1.3, lwd = 1.95)

# plot visual speeds equivalent to walking speeds
points(0.4166, 0.6944,  pch = 3, col = scales::alpha("gray50", alpha = 0.9), cex = 1.3, lwd = 1.95) # 0.4166 m/s (1.5 km/h) - 0.6944 m/s (2.5 km/h)
# export 4 x 4.5

# Slope coefficient 
lm(d$mF ~ d$mS - 1)
#abline(mod <- lm(d$mF ~ d$mS - 1)) # practically overlaps with the line for accurate ratio

```

<a name="S3"></a>
&nbsp;

#####**3. Compare matched optic flow speeds at different conditions** 

```{r}

#(1) Different walking speeds (slower / faster)

# summary statistics
describe(d$mS)
describe(d$mF)

# test distributions
shapiro.test(d$mS)
shapiro.test(d$mF)

# test differences
test <- wilcoxsign_test(d$mF ~ d$mS, paired = T, distribution = "exact")
test

# effect size
test@statistic@teststatistic / sqrt(nrow(d))

# correlation
cor.test(d$mS, d$mF, method = "spearman", exact = F)


#(2) Different starting speeds of visual flow (slower / faster)

# summary statistics
describe(d$m_starting_S)
describe(d$m_starting_F)

# test distributions
shapiro.test(d$m_starting_S)
shapiro.test(d$m_starting_F)

# test differences
test <- wilcoxsign_test(d$m_starting_F ~ d$m_starting_S, paired = T, distribution = "exact")
test

# effect size
test@statistic@teststatistic / sqrt(nrow(d))


#(3) Different colors of visualizations (green / red)

# summary statistics
describe(d$m_red)
describe(d$m_green)

# test distributions
shapiro.test(d$m_red)
shapiro.test(d$m_green)

# test differences
test <- wilcoxsign_test(d$m_green ~ d$m_red, paired = T, distribution = "exact")
test

# effect size
test@statistic@teststatistic / sqrt(nrow(d))

d_plot <- select(d, ID, diff_walking_speed, diff_starting_speed, diff_color)

d_plot <- gather(d_plot, key = conditions, value = estimate, -"ID")

d_plot$conditions <- as.factor(d_plot$conditions)
levels(d_plot$conditions)

d_plot$conditions <- ordered(d_plot$conditions, levels = c("diff_walking_speed", "diff_starting_speed", "diff_color"))

col1 <- "#d1ab02"
col2 <- darken("gray70", amount = 0.4)
col2 <- "#7a7696" # "#7a7696"

fig <- ggplot(d_plot, aes(x=factor(conditions), y=estimate, fill=factor(conditions)) ) +  geom_jitter(position=position_jitter(width=0.1, height=0.0), aes(colour=factor(conditions)), alpha=0.5, size = 1.7) + geom_boxplot(width = 0.45, alpha = 0.5, show.legend = FALSE, outlier.shape = NA) + scale_color_manual(values=c(col1, col2, col2)) + scale_fill_manual(values=c(col1, col2, col2)) + labs(x = "conditions" , y = "Difference between estimates (m/s)") + scale_y_continuous(limits= c(-0.39, 1.42), breaks = c(-0.4,-0.2,0,0.2, 0.4,0.6,0.8,1,1.2,1.4), expand = c(0.02,0.02)) + scale_x_discrete(labels = c('Walking speed','Starting visual speed', 'Colour')) + theme_classic() + theme(legend.position = "none", axis.text.y = element_text(face = "bold", size = 14.5, color = "black"), axis.text.x = element_text(face = "bold", size = 10, color = "black"), axis.title.y = element_text(margin = margin(t = 0, r = 9, b = 0, l = 0)), axis.title.x = element_text(margin = margin(t = 5, r = 0, b = 0, l = 0))) + geom_hline(yintercept=0, linetype="dashed", color = scales::alpha("black",0.5))

fig

```


<a name="S4"></a>
&nbsp;

#####**4. Explore associations bewteen optic flow estimates and confidence & anthropometric & proprioceptive data** 

```{r}

## Compare variance of estimates at different walking speeds

# summary statistics
describe(d$sdS)
describe(d$sdF)

# test distributions
shapiro.test(d$sdS)
shapiro.test(d$sdF)

# test differences
test <- wilcoxsign_test(d$sdF ~ d$sdS, paired = T, distribution = "exact")
test

## Describe confidence ratings

# Confidence ratings
describe(d$resp_confidence)

# Correlation between confidence ratings and variance estimates
shapiro.test(d$resp_confidence)
d$of_variance <- (d$sdF + d$sdS)/2
cor.test(d$resp_confidence, d$of_variance, method = "spearman", exact = F)

# Correlation between confidence ratings and mean speed of estimates
d$of_speed <- (d$mF + d$mS)/2
cor.test(d$resp_confidence, d$of_speed, method = "spearman", exact = F)

# Correlation between anthropomorphic variables and optic flow estimates
shapiro.test(d$of_speed)
shapiro.test(d$of_variance)

cor.test(d$bmi, d$of_speed, method = "spearman", exact = F)
cor.test(d$bmi, d$of_variance, method = "spearman", exact = F)

# suggestibility 
shapiro.test(d$diff_starting_speed)
cor.test(d$bmi, d$diff_starting_speed, method = "spearman", exact = F)
#scatter.smooth(d$bmi, d$diff_starting_speed)

cor.test(d$bmi, d$diff_walking_speed, method = "spearman", exact = F)
#scatter.smooth(d$bmi, d$diff_walking_speed)

cor.test(d$bmi, d$diff_color, method = "spearman", exact = F)
#scatter.smooth(d$bmi, d$color)

cor.test(d$bmi, d$resp_confidence, method = "spearman", exact = F)
#scatter.smooth(d$bmi, d$resp_confidence)

#write.table(d, "OPTVR_determination_of_optic_flow_speed_output", sep = ",")

# merge data frames
d <- merge(d, dt_walking, by = "ID")

# test associations beteween sensitivity and average optic flow speed
cor.test(d$s, d$of_speed, method = "spearman", exact = F)
#scatter.smooth(d$s, d$of_speed)

# test associations beteween sensitivity and variance of estimates
cor.test(d$s, d$of_variance, method = "spearman", exact = F)
#scatter.smooth(d$s, d$of_variance)

# correlation between confidence ratings between tasks
cor.test(d$resp_confidence, d$conf_acc_dec_mean, method = "spearman", exact = F)

# mean confidence from both tasks and its correlation with BMI
d$mean_conf <- (d$conf_acc_dec_mean + d$resp_confidence) / 2
shapiro.test(d$mean_conf)
cor.test(d$bmi, d$mean_conf, method = "pearson")

# optional: plot relation between these variables
bold.text <- element_text(face = "bold", color = "black")

fig <- ggplot(data = d, aes(x = bmi, y = resp_confidence)) + 
  geom_smooth(col = "grey8", method = "lm", level=0.95, alpha = 0.5) + 
  geom_point(col = "grey 15", alpha = 0.6, size = 3, shape = 19) + 
  labs(x = "Body Mass Index", y = "Mean confidence ratings") + 
 scale_fill_manual() +  theme_classic() +  theme(axis.text=element_text(size=15, colour = "black", face="bold"),  axis.title=element_text(size=17), axis.title.y = element_text(margin = margin(t = 0, r = 3, b = 0, l = 0)), axis.title.x =   element_text(margin = margin(t = 5, r = 0, b = 0, l = 0))) + theme(axis.title = bold.text) +
  scale_y_continuous(limits= c(min(d$mean_conf) - 0.5, max(d$mean_conf) + 0.5), expand = c(0.01,0.01)) + 
  scale_x_continuous(limits= c(15, max(d$bmi) + 0.5), expand = c(0.01,0.01)) +
  theme(axis.title = bold.text)
#fig  

```

